---
title: '2019_05_02_segmentation_controls'
author: "Toni Beltran"
date: "02/05/2019"
output: html_document
---

Segmentation analysis through HMM fitting and retrieval of the fitted chain of hidden states:
 - the time course data (two runs of 12 generations) is used to fit a 2-state HMM - initialized with the means and standard deviations of the     clusters obtained by k-means analysis
 - the internal states are recovered and analysed in the same way as in the k-means segmentation analysis
This analysis results in lower rates of epimutation as single generation changes are sometimes considered not a change - since the distribution of 22G counts for each hidden state is also a free parameter.


```{r load data}


library(gplots)
library(RColorBrewer)
library(ggplot2)
library(depmixS4)
library(DESeq2)

setwd("/Users/ab6415/Bioinformatics/Analysis/EpiMALs_PAPER_ANALYSES/05_Figure3/")

ttg_DEseqnorm<-read.table("../02_Normalised_counts/22G_DEseqnorm_counts_averaged.txt")
ttg_MAplot_data<-read.table("../03_Figure1/MAplot_filtering_22Gs_genbygen_p1e-2.txt",header=TRUE)
ttg_MAplot_data<-ttg_MAplot_data[which(ttg_MAplot_data$padj<1e-4),]

summary_stats_A<-data.frame(mean=apply(ttg_DEseqnorm[,1:12],MARGIN=1,FUN = mean),
                            min=apply(ttg_DEseqnorm[,1:12],MARGIN=1,FUN = min))

summary_stats_B<-data.frame(mean=apply(ttg_DEseqnorm[,13:24],MARGIN=1,FUN = mean),
                            min=apply(ttg_DEseqnorm[,13:24],MARGIN=1,FUN = min))

summary_stats_AB<-data.frame(mean=apply(ttg_DEseqnorm[,1:24],MARGIN = 1,FUN=mean),
                             min=apply(ttg_DEseqnorm[,1:24],MARGIN = 1,FUN=min))

```

```{r DEseq}
#DEseq to remove batch variation (early vs late, plus B1-B4 samples)

library(DESeq2)
ttg_A_rawcounts<-read.table("../01_Raw_counts/RAW_22G_COUNTS/22G_RAW_COUNTS_lineageA.txt")
ttg_B_rawcounts<-read.table("../01_Raw_counts/RAW_22G_COUNTS/22G_RAW_COUNTS_lineageB.txt")
ttg_AB_rawcounts<-merge(ttg_A_rawcounts,ttg_B_rawcounts,by=0); rownames(ttg_AB_rawcounts)<-ttg_AB_rawcounts$Row.names; ttg_AB_rawcounts$Row.names<-NULL

cds<-DESeqDataSetFromMatrix(countData = ttg_AB_rawcounts,
                            colData = data.frame(line=colnames(ttg_AB_rawcounts),condition=as.factor(rep(c(rep(1,7),rep(2,5)),2))),
                            design= ~condition)
dds<-DESeq(cds)
res<-results(dds)
res<-res[order(res$padj,decreasing = FALSE),]

ggplot(as.data.frame(res))+geom_histogram(aes(x=pvalue))
ggplot(as.data.frame(res))+geom_histogram(aes(x=padj))

removegenes_lastfive<-rownames(res[which(res$padj<0.2 & abs(res$log2FoldChange)>1),])


cds<-DESeqDataSetFromMatrix(countData = ttg_AB_rawcounts,
                            colData = data.frame(line=colnames(ttg_AB_rawcounts),condition=as.factor(c(rep(1,12),2,rep(1,3),2,rep(1,7)))),
                            design= ~condition)
dds<-DESeq(cds)
res<-results(dds)
res<-res[order(res$padj,decreasing = FALSE),]

ggplot(as.data.frame(res))+geom_histogram(aes(x=pvalue))
ggplot(as.data.frame(res))+geom_histogram(aes(x=padj))

removegenes_b1_b5<-rownames(res[which(res$padj<0.2 & abs(res$log2FoldChange)>1),])

removegenes<-union(removegenes_lastfive,removegenes_b1_b5)
length(removegenes)



epimutable_genes_genbygen<-ttg_MAplot_data$ID[-which(ttg_MAplot_data$ID %in% removegenes)]
length(unique(epimutable_genes_genbygen))

ttg_DEseqnorm_epimutable_genes<-ttg_DEseqnorm[which(rownames(ttg_DEseqnorm) %in% epimutable_genes_genbygen),]
nrow(ttg_DEseqnorm_epimutable_genes)

```

```{r HMM segmentation}

#initialise HMM with prior information

kmeans_row<-function(vector){  
  kmc<-kmeans(x = as.numeric(vector),centers=2)
  high<-kmc$cluster[which.is.max(vector)]
  if (high==2){return(kmc$cluster)
  }else{return(abs(kmc$cluster-3))}
}

hmm_fit_and_estimate_underlying_states_2stHMM_kmeansinit<-function(trace){
  
  kmeans_seg<-kmeans_row(trace)
  mean_st1<-mean(trace[which(kmeans_seg==1)])
  sd_st1<-0.1*mean_st1
  mean_st2<-mean(trace[which(kmeans_seg==2)])
  sd_st2<-0.1*mean_st2
  
  respstart<-c(c(mean_st1,sd_st1),c(mean_st2,sd_st2))
  
  trace<-trace[1:24]
  dat<-data.frame(trace=trace)
  mod <- depmix(trace ~ 1, family = gaussian(), nstates = 2, data = dat, respstart = respstart, ntimes = c(12,12))
  fm2 <- tryCatch(fit(mod, verbose = FALSE), error=function(e) 1)

 #calculate trajectory of hidden states
  final_model_states<-tryCatch(posterior(fm2), error=function(e) 1)
  
  high<-tryCatch(final_model_states$state[which.is.max(trace)], error=function(e) 0)
  if (high==2){return(final_model_states$state)
  }else if (high==1){return(abs(final_model_states$state-3))}
  else{return(rep(0,24))}

}


hmm_segmentation<-matrix(0,nrow=nrow(ttg_DEseqnorm_epimutable_genes),ncol = 24)
rownames(hmm_segmentation)<-rownames(ttg_DEseqnorm_epimutable_genes)
for (gene in rownames(ttg_DEseqnorm_epimutable_genes)){
      hmm_segmentation[gene,]<-hmm_fit_and_estimate_underlying_states_2stHMM_kmeansinit(as.numeric(ttg_DEseqnorm_epimutable_genes[gene,]))
}

hmm_segmentation<-hmm_segmentation[-which(apply(hmm_segmentation,MARGIN=1,FUN=sum)==0),]
colnames(hmm_segmentation)<-colnames(ttg_DEseqnorm_epimutable_genes)[1:24]


epimuts_from_hmm_trace<-function(vect,gene){
  trace<-rle(vect)
  epimut_number<-length(trace$lengths)
  completeness<-rep(1,epimut_number); completeness[1]<-0; completeness[epimut_number]<-0; 
  return(data.frame(ID=rep(gene,epimut_number),up_or_down=trace$values,duration=trace$lengths,completeness=completeness,stgen=c(-1,cumsum(trace$lengths)[1:(epimut_number-1)])))
}


epimut_duration_hmm_A<-data.frame()
epimut_duration_hmm_B<-data.frame()

for (gene in rownames(hmm_segmentation)){
  epimut_duration_hmm_A<-rbind(epimut_duration_hmm_A,epimuts_from_hmm_trace(hmm_segmentation[gene,1:12],gene))
  epimut_duration_hmm_B<-rbind(epimut_duration_hmm_B,epimuts_from_hmm_trace(hmm_segmentation[gene,13:24],gene))

}

epimut_duration_hmm_A<-epimut_duration_hmm_A[-which(epimut_duration_hmm_A$stgen==(-1)),]
epimut_duration_hmm_A<-epimut_duration_hmm_A[-which(epimut_duration_hmm_A$stgen==(12)),]

epimut_duration_hmm_B<-epimut_duration_hmm_B[-which(epimut_duration_hmm_B$stgen==(-1)),]
epimut_duration_hmm_B<-epimut_duration_hmm_B[-which(epimut_duration_hmm_B$stgen==(12)),]


table(table(epimut_duration_hmm_A$ID))
table(table(epimut_duration_hmm_B$ID))


remove_A<-names(which(table(epimut_duration_hmm_A$ID)>7))
remove_B<-names(which(table(epimut_duration_hmm_B$ID)>7))

epimut_duration_hmm_A<-epimut_duration_hmm_A[-which(epimut_duration_hmm_A$ID %in% remove_A),]
epimut_duration_hmm_B<-epimut_duration_hmm_B[-which(epimut_duration_hmm_B$ID %in% remove_B),]


```

```{r correct for missing generation 7}


#correction to incorporate missing generation 7
#assumes no change from generation 6 to 7 (overestimates duration)
epimut_duration_hmm_A$stgen_corrected<-epimut_duration_hmm_A$stgen
epimut_duration_hmm_A$duration_corrected<-epimut_duration_hmm_A$duration
for (i in seq(nrow(epimut_duration_hmm_A))){
  if (epimut_duration_hmm_A[i,"stgen"]>6){
  epimut_duration_hmm_A[i,"stgen_corrected"]<-epimut_duration_hmm_A[i,"stgen_corrected"]+1
  }
  if ((epimut_duration_hmm_A[i,"stgen"]+epimut_duration_hmm_A[i,"duration"])>6 & epimut_duration_hmm_A[i,"stgen"]<7){
  epimut_duration_hmm_A[i,"duration_corrected"]<-epimut_duration_hmm_A[i,"duration_corrected"]+1
  }
}


epimut_duration_hmm_B$stgen_corrected<-epimut_duration_hmm_B$stgen
epimut_duration_hmm_B$duration_corrected<-epimut_duration_hmm_B$duration
for (i in seq(nrow(epimut_duration_hmm_B))){
  if (epimut_duration_hmm_B[i,"stgen"]>6){
  epimut_duration_hmm_B[i,"stgen_corrected"]<-epimut_duration_hmm_B[i,"stgen_corrected"]+1
  }
  if ((epimut_duration_hmm_B[i,"stgen"]+epimut_duration_hmm_B[i,"duration"])>6 & epimut_duration_hmm_B[i,"stgen"]<7){
  epimut_duration_hmm_B[i,"duration_corrected"]<-epimut_duration_hmm_B[i,"duration_corrected"]+1
  }
}

epimut_duration_hmm_A$duration<-epimut_duration_hmm_A$duration_corrected
epimut_duration_hmm_B$duration<-epimut_duration_hmm_B$duration_corrected
epimut_duration_hmm_A$stgen<-epimut_duration_hmm_A$stgen_corrected
epimut_duration_hmm_B$stgen<-epimut_duration_hmm_B$stgen_corrected

epimut_duration_hmm_A$stgen_corrected<-NULL; epimut_duration_hmm_B$stgen_corrected<-NULL
epimut_duration_hmm_A$duration_corrected<-NULL; epimut_duration_hmm_B$duration_corrected<-NULL



```


```{r plot data}

##all epimutations
table(epimut_duration_hmm_A$duration)
table(epimut_duration_hmm_A$up_or_down)
table(epimut_duration_hmm_A[which(epimut_duration_hmm_A$completeness==0),"duration"])
table(epimut_duration_hmm_A[which(epimut_duration_hmm_A$completeness==1),"duration"])

A_data_for_barplot<-data.frame(duration=rep(seq(1,11),2),
                               count=c(table(factor(epimut_duration_hmm_A[which(epimut_duration_hmm_A$completeness==1),"duration"],levels=c(1:11))),
                                       table(factor(epimut_duration_hmm_A[which(epimut_duration_hmm_A$completeness==0),"duration"],levels=c(1:11)))),
                               completeness=c(rep("complete",11),rep("incomplete",11)))

# Grouped
ggplot(A_data_for_barplot, aes(fill=completeness, y=count, x=duration)) + 
    geom_bar(position="dodge", stat="identity") + ggtitle("lineage A, hmm segmentation")+theme_classic()
ggsave("../05_Figure3/HMM_segmentation_duration_barplots_A.pdf",dpi="retina")



##all epimutations
table(epimut_duration_hmm_B$duration)
table(epimut_duration_hmm_B$up_or_down)
table(epimut_duration_hmm_B[which(epimut_duration_hmm_B$completeness==0),"duration"])
table(epimut_duration_hmm_B[which(epimut_duration_hmm_B$completeness==1),"duration"])

B_data_for_barplot<-data.frame(duration=rep(seq(1,11),2),
                               count=c(table(factor(epimut_duration_hmm_B[which(epimut_duration_hmm_B$completeness==1),"duration"],levels=c(1:11))),
                                       table(factor(epimut_duration_hmm_B[which(epimut_duration_hmm_B$completeness==0),"duration"],levels=c(1:11)))),
                               completeness=c(rep("complete",11),rep("incomplete",11)))

# Grouped
ggplot(B_data_for_barplot, aes(fill=completeness, y=count, x=duration)) + 
    geom_bar(position="dodge", stat="identity") + ggtitle("lineage B, hmm segmentation")+theme_classic()
ggsave("../05_Figure3/HMM_segmentation_duration_barplots_B.pdf",dpi="retina")


#both lineages

##all epimutations
epimut_duration_hmm_both<-rbind(epimut_duration_hmm_A,epimut_duration_hmm_B)

table(epimut_duration_hmm_both$duration)
table(epimut_duration_hmm_both$up_or_down)
table(epimut_duration_hmm_both[which(epimut_duration_hmm_both$completeness==0),"duration"])
table(epimut_duration_hmm_both[which(epimut_duration_hmm_both$completeness==1),"duration"])

both_data_for_barplot<-data.frame(duration=rep(seq(1,11),2),
                               count=c(table(factor(epimut_duration_hmm_both[which(epimut_duration_hmm_both$completeness==1),"duration"],levels=c(1:11))),
                                       table(factor(epimut_duration_hmm_both[which(epimut_duration_hmm_both$completeness==0),"duration"],levels=c(1:11)))),
                               completeness=c(rep("complete",11),rep("incomplete",11)))

# Grouped
ggplot(both_data_for_barplot, aes(fill=completeness, y=count, x=duration)) + 
    geom_bar(position="dodge", stat="identity") + ggtitle("both lineages, hmm segmentation")+theme_classic()
ggsave("../05_Figure3/HMM_segmentation_duration_barplots_bothAB.pdf",dpi="retina")


write.table(epimut_duration_hmm_A,file="../05_Figure3/epimut_duration_batchremoved_HMM_linA_JOINT_p1e-4.txt",quote = FALSE,row.names = FALSE,sep="\t")
write.table(epimut_duration_hmm_B,file="../05_Figure3/epimut_duration_batchremoved_HMM_linB_JOINT_p1e-4.txt",quote = FALSE,row.names = FALSE, sep="\t")


```







