---
title: "2019_04_26_correlation_analyses_genbygen.Rmd"
author: "Toni Beltran"
date: "26/04/2019"
output: html_document
---

This code assesses the separation between samples as a function of the number of generations separating them.

```{r load data}

library(ggplot2)
library(MASS)
library(viridis)
library(gplots)

setwd("/Users/ab6415/Documents/Work_laptop_stuff/Bioinformatics/Analysis/EpiMALs_PAPER_ANALYSES_Dryad/05_Figure3/")
ttg_DEseqnorm<-read.table("../02_Normalised_counts/22G_DEseqnorm_counts_averaged.txt")


```


```{r define cor function}

#this function calculates the all vs all correlations between samples for a particular set of genes (the "gene_set" input variable), and plots them as a function of the number of generations separating them.

plot_allvsall_cors_twoLineages<-function(df,gene_set,title,plot){
  df<-df[which(rownames(df) %in% gene_set),]
  
  if (plot==TRUE){
  heatmap.2(cor(as.matrix(df)),Rowv = FALSE,Colv = FALSE,trace="none",main=title)
  heatmap.2(cor(as.matrix(df)),trace="none",main=title)
  }
  
  gendistances<-read.table("genbygen_distances_2lineages.txt"); gendistances<-gendistances[-c(7,20),-c(7,20)]
  colnames(gendistances)<-colnames(df)

  final_cordata<-data.frame(cors="",gens="")
  final_cordata$cors<-as.numeric(final_cordata$cors)
  final_cordata$gens<-as.numeric(final_cordata$gens)
  
  for (i in seq(24)){
    cordata<-data.frame(cors=cor(as.matrix(df))[which(gendistances==i)],gens=rep(i,length(which(gendistances==i))))
    final_cordata<-rbind.data.frame(final_cordata,cordata)
}

final_cordata<-final_cordata[-1,]
final_cordata$gens<-as.factor(final_cordata$gens)

if (plot==TRUE){
print(ggplot(final_cordata)+geom_dotplot(aes(x=final_cordata$gens,y=final_cordata$cors),binaxis='y', stackdir='center')+xlab("number of generations separating samples")+ylab("Pearson correlation coefficient"))+ggtitle(title)

print(ggplot(final_cordata)+geom_boxplot(aes(x=final_cordata$gens,y=final_cordata$cors))+xlab("number of generations separating samples")+ylab("Pearson correlation coefficient"))+ggtitle(title)
}
if (plot==TRUE){
print(cor(as.numeric(final_cordata$gens),as.numeric(final_cordata$cors)))
print(cor.test(as.numeric(final_cordata$gens),as.numeric(final_cordata$cors)))
}
test<-cor.test(as.numeric(final_cordata$gens),as.numeric(final_cordata$cors))
return(cor(as.numeric(final_cordata$gens),as.numeric(final_cordata$cors)))
  
}

get_onelin_genes<-function(setA,setB){
  setA_alone<-setA[-which(setA %in% setB)]
  setB_alone<-setB[-which(setB %in% setA)]
  return(union(setA_alone,setB_alone))
}


```

This part of the code removes the genes contributing to the batch effect observed between generations 1-8 and 9-13, as well as the genes contributing to the B1 and B4 outlier samples.

```{r DEseq 1to8 vs 9to13 - remove batch genes}

library(DESeq2)
ttg_A_rawcounts<-read.table("../01_Raw_counts/RAW_22G_COUNTS/22G_RAW_COUNTS_lineageA.txt")
ttg_B_rawcounts<-read.table("../01_Raw_counts/RAW_22G_COUNTS/22G_RAW_COUNTS_lineageB.txt")
ttg_AB_rawcounts<-merge(ttg_A_rawcounts,ttg_B_rawcounts,by=0); rownames(ttg_AB_rawcounts)<-ttg_AB_rawcounts$Row.names; ttg_AB_rawcounts$Row.names<-NULL

cds<-DESeqDataSetFromMatrix(countData = ttg_AB_rawcounts,
                            colData = data.frame(line=colnames(ttg_AB_rawcounts),condition=as.factor(rep(c(rep(1,7),rep(2,5)),2))),
                            design= ~condition)
dds<-DESeq(cds)
res<-results(dds)
res<-res[order(res$padj,decreasing = FALSE),]

hist(res$pvalue)
hist(res$padj)

nrow(res[which(res$padj<0.05 & abs(res$log2FoldChange)>1),])
nrow(res[which(res$padj<0.1 & abs(res$log2FoldChange)>1),])
nrow(res[which(res$padj<0.2 & abs(res$log2FoldChange)>1),])
removegenes_lastfive<-rownames(res[which(res$padj<0.2 & abs(res$log2FoldChange)>1),])


cds<-DESeqDataSetFromMatrix(countData = ttg_AB_rawcounts,
                            colData = data.frame(line=colnames(ttg_AB_rawcounts),condition=as.factor(c(rep(1,12),2,rep(1,3),2,rep(1,7)))),
                            design= ~condition)
dds<-DESeq(cds)
res<-results(dds)
res<-res[order(res$padj,decreasing = FALSE),]

hist(res$pvalue)
hist(res$padj)

nrow(res[which(res$padj<0.05 & abs(res$log2FoldChange)>1),])
nrow(res[which(res$padj<0.1 & abs(res$log2FoldChange)>1),])
nrow(res[which(res$padj<0.2 & abs(res$log2FoldChange)>1),])
removegenes_b1_b5<-rownames(res[which(res$padj<0.2 & abs(res$log2FoldChange)>1),])

removebatch<-function(setA,setB){
  #setA_alone<-setA[-which(setA %in% setB)]
  #setB_alone<-setB[-which(setB %in% setA)]
  unionAB<-union(setA,setB)
  print(length(unionAB))
  unionAB<-unionAB[-which(unionAB %in% removegenes)]
  print(length(unionAB))
  return(unionAB)
}


removegenes<-union(removegenes_lastfive,removegenes_b1_b5)

```


# Correlation analysis for epimutable genes

```{r plot correlations MAplot filtered genes}

ttg_MAplot_data<-read.table("../03_04_Figures1-2/MAplot_filtering_22Gs_genbygen_p1e-2.txt",header=TRUE)
ttg_MAplot_data<-ttg_MAplot_data[-which(ttg_MAplot_data$ID %in% removegenes),]
ttg_MAplot_data<-ttg_MAplot_data[which(ttg_MAplot_data$padj<1e-4),]
length(unique(ttg_MAplot_data$ID))

plot_allvsall_cors_twoLineages(log2(ttg_DEseqnorm[,1:24]+1),ttg_MAplot_data$ID,"epimutated genes",plot=TRUE)


```

## Correlation analysis for CSR-1 target genes

The scaling of the correlation is not observed for CSR-1 targets. It is observed for HRDE-1 targets to a lesser extent than for the epimutations set (see also bootstrap analysis of groups of 145 genes below to control for gene set size).

```{r other sets of genes that should not have this correlation}

csr1_targets<-read.table("../07_Gene_Sets/CSR1_targets"); csr1_targets<-csr1_targets$x
csr1_targets_nobatch<-removebatch(csr1_targets,csr1_targets)

hrde1_targets<-read.table("../07_Gene_Sets/WAGO9_targets"); hrde1_targets<-hrde1_targets$x
hrde1_targets_nobatch<-removebatch(hrde1_targets,hrde1_targets)

plot_allvsall_cors_twoLineages(log2(ttg_DEseqnorm[,1:24]+1),csr1_targets_nobatch,"csr-1 targets",plot=TRUE)
plot_allvsall_cors_twoLineages(log2(ttg_DEseqnorm[,1:24]+1),hrde1_targets_nobatch,"hrde-1 targets",plot=TRUE)



```

## Bootstrap analysis of the correlation coefficients

Sets of 145 csr-1 targets are sampled and the correlation coefficient describing the relationship between similarity and distance in generations is calculated for each case - the resulting distribution is compared to the correlation coefficient observed for the set of 145 epimutable genes. The same analysis is applied to HRDE-1 targets.

```{r csr-1 and hrde-1 bootstrapping analysis}

length(unique(ttg_MAplot_data$ID))
cors_csr1<-rep(0,1000)
for (i in seq(1000)){
  #print(i)
  csr_1_subset<-sample(csr1_targets_nobatch,145)
  cors_csr1[i]<-plot_allvsall_cors_twoLineages(log2(ttg_DEseqnorm[,1:24]+1),csr_1_subset,"csr-1 subsets",plot=FALSE)
}
csr_1_cordata<-data.frame(cors_csr1=cors_csr1)
ggplot(csr_1_cordata)+geom_histogram(aes(x=cors_csr1),breaks=seq(-0.8,0.2,0.025))+ggtitle("CSR-1 bootstrapped correlation coefficients")+geom_vline(xintercept=-0.7739395,color="red")



cors_hrde1<-rep(0,1000)
for (i in seq(1000)){
  #print(i)
  hrde_1_subset<-sample(hrde1_targets_nobatch,145)
  cors_hrde1[i]<-plot_allvsall_cors_twoLineages(log2(ttg_DEseqnorm[,1:24]+1),hrde_1_subset,"hrde-1 subsets",plot=FALSE)
}
hrde_1_cordata<-data.frame(cors_hrde1=cors_hrde1)
ggplot(hrde_1_cordata)+geom_histogram(aes(x=cors_hrde1),breaks=seq(-0.8,0.2,0.025))+ggtitle("HRDE-1 bootstrapped correlation coefficients")+geom_vline(xintercept=-0.7739395,color="red")

```

## Plotting correlation scalings for CSR-1 genes and epimutated genes together


```{r plot csr1 and ma correlations together}

plot_allvsall_cors_twoLineages_return_cordata<-function(df,gene_set,title){
  df<-df[which(rownames(df) %in% gene_set),]

  gendistances<-read.table("genbygen_distances_2lineages.txt"); gendistances<-gendistances[-c(7,20),-c(7,20)]
  colnames(gendistances)<-colnames(df)

  final_cordata<-data.frame(cors="",gens="")
  final_cordata$cors<-as.numeric(final_cordata$cors)
  final_cordata$gens<-as.numeric(final_cordata$gens)
  
  for (i in seq(26)){
    cordata<-data.frame(cors=cor(as.matrix(df))[which(gendistances==i)],gens=rep(i,length(which(gendistances==i))))
    final_cordata<-rbind.data.frame(final_cordata,cordata)
}

final_cordata<-final_cordata[-1,]
nrow(final_cordata)
final_cordata$gens<-as.factor(final_cordata$gens)

return(final_cordata)
  
}


ma_fdr0.01_cordata<-plot_allvsall_cors_twoLineages_return_cordata(log2(ttg_DEseqnorm[,1:24]+1),unique(ttg_MAplot_data$ID),"tit")
csr1_cordata<-plot_allvsall_cors_twoLineages_return_cordata(log2(ttg_DEseqnorm[,1:24]+1),csr1_targets_nobatch,"tit")
ma_fdr0.01_cordata$group<-rep("ma",nrow(ma_fdr0.01_cordata))
csr1_cordata$group<-rep("csr1",nrow(csr1_cordata))

cordata_combined<-rbind(ma_fdr0.01_cordata,csr1_cordata)


ggplot(cordata_combined,aes(x=gens,y=cors))+geom_dotplot(aes(fill=group,color=group),binaxis='y',stackdir='center',binwidth=0.003,alpha=0.8)+
  xlab("number of generations separating samples")+
  ylab("Pearson correlation coefficient")+
  stat_summary(data=cordata_combined[which(cordata_combined$group=="ma"),],mapping=aes(x=gens,y=cors),fun.y = median, fun.ymin = median, fun.ymax = median,geom = "point", width = 0.5,color="black")+
  stat_summary(data=cordata_combined[which(cordata_combined$group=="csr1"),],mapping=aes(x=gens,y=cors),fun.y = median, fun.ymin = median, fun.ymax = median,geom = "point", width = 0.5,color="grey")+
  theme_classic()
ggsave("epimutated_genes_vs_CSR1_targets_cors_dotplot.pdf",dpi="retina",width=9)



fit_csr<-lm(data = csr1_cordata,formula = cors~as.numeric(gens))
summary(fit_csr)
fit_ma<-lm(data = ma_fdr0.01_cordata,formula = cors~as.numeric(gens))
summary(fit_ma)

ggplot(cordata_combined,aes(x=gens,y=cors))+
  geom_point(aes(fill=group,color=group),binaxis='y', stackdir='center',binwidth=0.003,alpha=0.8)+
  xlab("number of generations separating samples")+
  ylab("Pearson correlation coefficient")+
  geom_abline(slope=fit_ma$coefficients[2],intercept = fit_ma$coefficients[1],color="red")+
  geom_abline(slope=fit_csr$coefficients[2],intercept = fit_csr$coefficients[1],color="blue")+
stat_summary(data=cordata_combined[which(cordata_combined$group=="ma"),],mapping=aes(x=gens,y=cors),fun.y = median, fun.ymin = median, fun.ymax = median,geom = "crossbar", width = 0.5,color="red")+
stat_summary(data=cordata_combined[which(cordata_combined$group=="csr1"),],mapping=aes(x=gens,y=cors),fun.y = median, fun.ymin = median, fun.ymax = median,geom = "crossbar", width = 0.5,color="black")+
  theme_classic()
ggsave("epimutated_genes_vs_CSR1_targets_cors_dotplot_with_linefit.pdf")

cor.test(fit_csr$model$cors,fit_csr$model$`as.numeric(gens)`,method = "pearson")
cor.test(fit_ma$model$cors,fit_ma$model$`as.numeric(gens)`,method = "pearson")




```

